\LARGE \bf Expectation-Maximization based defense mechanism for distributed Model Predictive Control

### 2022-02-08 article/article.tex
simulation, figures, table and some text

\emph{Notation:} In this paper, $\norm{\cdot}$ and $\norm{\cdot}_{F}$ represent the $\ell_{2}$ and Frobenius norms. $\norm{\vec{v}}_{Y}$ is the weighted norm $\norm{Y^{\frac{1}{2}}\vec{v}}$.
$\Proj^{\set{T}}(\cdot)$ is the Euclidean projection onto set $\set{T}$.
$\card{x}$ is the number of elements in $x$.
$\kron{}{}$ represents the Kronecker product.
$\vectorize{A}$ vectorizes matrix $A$.
$\hadamard{}{}$ represents the Hadamard product.
$n\mathbin{:}i\mathbin{:}j$ is a row vector builder with elements $\{n,n+i,\dots,n+mi\}$, where $m=\mathrm{truncate}(\frac{j-n}{i})$, and ${n\mathbin{:}j}$ is equal to ${n\mathbin{:}1\mathbin{:}j}$.
$[A,B]$ and $[A;B]$ are horizontal and vertical concatenation of matrices of adequate sizes.
$0_{m\times n}$ and $1_{m\times n}$ are ${m\times n}$ matrices filled with $0$ and $1$.
$\0$ and $\1$ are 0 and 1 filled vectors of adequate size.
$\indicator{x}$ is the indicator function returning $1$ if $x$ is true and $0$ otherwise.
$\preceq$ is the component-wise less or equal for vectors and matrix inequality for matrices in ${\symmetric=\setbuild{X:\R^{n\times n}}{X'=X}}$.
${\defpos=\setbuild{X:\symmetric}{X\succ 0}}$, ${\semidefpos=\setbuild{X:\symmetric}{X\succeq 0}}$.
$I_{c}$ is the identity matrix of size $c$.
$A^{\dagger}$ is the generalized inverse ${{(A^{T}A)}^{-1}A^{T}}$.
$|A|$ is the determinant of $A$.
A vector $\vec{v}_{i}$, correspond to the $i$-th agent, and these vectors can be stacked in a vector $\vec{v}$.
$\random{a},\random{A},\randomvec{a}$ are random scalar, matrix and vector.
$\expectation{\random{x}}$ is the expected value of $\random{x}$, and $\probability{A \mid B}$ is the conditional probability of $A$ given condition $B$.
${\diag(A_{1},\dots,A_{N})}$ corresponds to a block diagonal matrix.

\section{PRELIMINARIES AND PROBLEM STATEMENT}\label{sec:PS}
\begin{equation}
\begin{matrix}
  \label{eq:systems}
\vec{x}_{i}[k+1]=A_{i}\xik + B_{i}\uik
\end{matrix}
\end{equation}

\begin{equation}
  \label{eq:upositive}
  \uik\succeq\0
\end{equation}

\begin{equation}
  \label{eq:constraint}
  \sum^{\nsubsystems}_{i=1}\Gamma_{i}\uik\preceq\umax
\end{equation}
where ${\Gamma_{i}:\R^{n_{u}\times n_{u}}}$\todo{$\Gamma_{i}:\semidefpos$?}, ${\umax:\R^{n_{u}\times 1}}$ and ${n_{u}=\card{\uik}}$.

\begin{problem}{Global \mpc{} Problem.}\label{Pb:GOP}
\begin{equation*}
\begin{matrix}
\underset{\useq}{\mathrm{minimize}}&\resizebox{0.35\textwidth}{!}{$\overbrace{\sum\limits^{M}_{i=1} \overbrace{\sum_{j=1}^{\predhorz} \norm{\mpcvec{v}[i][k+j][k]}^{2}_{Q_i}+\norm{\mpcvec{u}[i][k+j-1][k]}^{2}_{R_i}}^{\textstyle{} \obji}}^{\textstyle{} \globobj}$}\\
\mathrm{subject~ to}&~\eqref{eq:systems},\eqref{eq:upositive}\ \mathrm{and}\ \eqref{eq:constraint}
\left\}\small
\begin{aligned}
  &\forall i\in \{1:\nsubsystems\}\\
  &\forall j\in \{1:\predhorz\}
\end{aligned}\right.

\end{matrix}
% \label{eq:dmpcModLOP}
\end{equation*}
where ${Q_{i}:\semidefpos}$ and ${R_{i}:\defpos}$, $\vik$ is a control objective.
Reference tracking can be written as ${\vik=\wik-\xik}$, where $\wik$ is a state reference.
Disturbance rejection can be written as ${\vik=\xik}$.

$\optglobobj$ denotes the optimal value of the problem~\ref{Pb:GOP}, and ${\optuseq}$ is the optimal control sequences for all agents.
The problem~\ref{Pb:GOP} is solved at each time $k$, and the ${\optuikk}$ are applied in each respective $i$ subsystem, following a receding horizon strategy.
\end{problem}

\subsection{Distributed Model Predictive Control}\label{ssec:dMPC}
\todo{\dmpc{},
}
\begin{subequations}
  \begin{equation}
    \left.
      \small
      \begin{aligned}
        J_{i}^{\star}(\thetaik)&=\underset{\uiseq}{\mathrm{minimize}} \obji\\
        \mathrm{s.t.} &\quad\eqref{eq:systems}\ \&\ \eqref{eq:upositive}\\
        &\quad\Gamma_{i}\uik\preceq\thetaik:\lambdaik\\
      \end{aligned}
    \right\}
    \small
    \begin{aligned}
      &\forall i\in \{1:\nsubsystems\}\\
      &\forall j\in \{1:\predhorz\}
    \end{aligned}
    \label{eq:DOP_local}
  \end{equation}
  \begin{equation}
    \small
    \begin{aligned}
      J^{\star}&=\underset{\thetaseq}{\mathrm{minimize}}\sum^{\nsubsystems}_{i=1} J^{\star}_i(\thetaik)\\
      \mathrm{s.t.} &\quad \sum_{i=1}^{\nsubsystems}\thetaik\preceq\umax
    \end{aligned}
    \label{eq:DOP_master}
  \end{equation}
\end{subequations}

@@

Where ${\set{H} = \setbuild{\vec{\theta}}{\sum_{i=1}^{M}\vec{\theta}_{i}=\vec{u}_{\max}}}$, $\vec{\theta}$ are the allocations for all agents.
\todo{The sum $\sum_{i=1}^{M}\vec{\theta}_{i}$ can also be represented by the matrix multiplication $I_{c}^{M}\vec{\theta}$. Where ${c=N_{p}\pi_{\uik}}$.}

@@

where ${I_{c}^{M}=\kron{\1_{M,1}}{I_{c}}}$.
\todo{Observe that ${I_{c}^{M} \pseudoinv{I_{c}^{M}}\vec{v}=\frac{\sum_{i=0}^{M}\vec{v}_{i}}{M}}$}

\todo{
  But as the coordinator must also assure that ${\thetaik\succeq\0}$ so the local problems are feasible, we set 0 as lower bounds:
  }
  \todo[verify if need to use max]{
\begin{equation}
  % \label{eq:thetaNegot}
  \thetai\pplusone=
  \left\{ \thetai\p+\rho\left(\vec{\lambda}_{i}^{\star} (\vec{\theta}_{i}\p)-I_{c}^{M} \pseudoinv{I_{c}^{M}}\vec{\lambda}^{\star} (\vec{\theta}\p)\right)
  ,0\right\}
\end{equation}
}
\no{\todo[delete probably]{
  But observe in~\eqref{eq:projectedSubgradient} that if ${\vec{\theta}_{i}\p=\0}$, ${(\vec{\lambda}_{i}^{\star} (\vec{\theta}_{i}\p)-I_{c}^{M} \pseudoinv{I_{c}^{M}}\vec{\lambda}^{\star} (\vec{\theta}\p))}$ needs to be negative so $\vec{\theta}_{i}\pplusone$ becomes negative, which is impossible \question{is it though?} since $\lambdaik\succeq\0$, from the KKT optimality conditions, seen in~\cite[Chapter 5]{BoydVandenberghe2004}.
}}

@@

\subsection{Quadratic Case --- Formal Analysis}\label{ssec:FA}
% \todo{As seen in \cite{NogueiraEtAl2021}},

@@

If we take reference tracking, for instance, we have:
\begin{equation}
\small\begin{matrix*}[l]
 H_i&=&\mathcal{D}_i^T\bar{Q}_i\mathcal{D}_i+\bar{R}_i\\
\fik&=&\mathcal{D}_i^T\bar{Q}_i(\mathcal{M}_i\xik-\Wik)\\

@@

We stack in vectors $\Uik$ and $\Wik$
the input and setpoint predictions sequences ${\uiseq}$ and ${\wiseq}$.
${\mathcal{M}_{i}}$  and ${\mathcal{D}_{i}}$ are the prediction matrices of the \mpc{}.
$\bar{Q}_{i}$, $\bar{R}_{i}$, and $\Theta_{i}$ are block diagonal matrices built repeating $\predhorz$ times $Q_{i}$, $R_{i}$, and $\Gamma_i$ respectively.
\todo{
  Observe that $H_{i}:\defpos$.
}

\todo{As seen in~\cite{BemporadEtAl2002} and in~\cite{BorrelliEtAl2017}, the solution for a constrained linear quadratic regulation problem yields in a \pwa{} function.
Similarly, we can get an explicit  solution for its dual variables $\vec{\lambda}_{i}$, which is also a \pwa{} function with respect to $\vec{\theta}_{i}$:}
\begin{equation}
  \begin{aligned}
    \label{eq:lambdafuntheta}
    \lambdai=
    \begin{cases}
      -P_{i}^{1}\vec{\theta}_{i}-\vec{s}_{i}^{1}[k]&\text{if}\ G_{1}[k]\thetai \preceq z_{1}[k] \\
      \qquad\quad \vdots&\qquad\quad \vdots\\
      -P_{i}^{N}\vec{\theta}_{i}-\vec{s}_{i}^{N}[k]&\text{if}\ G_{N}[k]\thetai \preceq z_{N}[k] \\
    \end{cases}
  \end{aligned}
\end{equation}
The halfspaces defined by the set of pairs $(G_{j}[k],z_{j}[k])$ represent a combination of active constraints for a given time $k$.

\begin{challenge}\label{ch:partition_unknown}
  It is important to note that since the actual state is unknown we cannot anticipate the partition of the space.
\end{challenge}
\begin{challenge}\label{ch:zone_unknown}
 For the same reason the values of $\vec{s}_{i}[k]$ are also unknown.
\end{challenge}

 As each constraint can be active or inactive, for a group of $n$ constraints, we can have at most $N=2^{n}$ different combinations of active/inactive sets and consequently $N$ regions (or zones).
 \begin{remark}\label{re:N_different_zones}
  Observe that depending on the constraints, some of these combinations can redundant or even make the optimization infeasible.
 \end{remark}
 \begin{assumption}
   We assume that none of these cases in Remark~\ref{re:N_different_zones} can occur, and we have $N$ zones.
 \end{assumption}

@@

\section{Attack in DMPC scheme}\label{sec:attack}

@@

@@

A conclusion section is not required. Although a conclusion may review
the main points of the paper, do not replicate the abstract as the
conclusion. A conclusion might elaborate on the importance of the work
or suggest applications and extensions.
\begin{ack}

\end{ack}

>>>>

\emph{Notation:}
In this paper, $\norm{\cdot}$, $\norm{\cdot}_{F}$, and $\norm{\vec{v}}_{Y}$ represent the $\ell_{2}$ norm, the Frobenius norm and the weighted norm $\norm{Y^{\frac{1}{2}}\vec{v}}$.
$\Proj^{\set{T}}(\cdot)$ is the Euclidean projection onto set $\set{T}$.
$\card{x}$ is the number of elements in $x$.
$\kron{}{}$, and $\hadamard{}{}$ are the Kronecker and Hadamard products.
$\vectorize{A}$ vectorizes matrix $A$.
${n\mathbin{:}i\mathbin{:}j}$ is a row vector builder with elements $\{n,n+i,\dots,n+mi\}$, where ${m=\mathrm{truncate}(\frac{j-n}{i})}$, and ${n\mathbin{:}j}$ is equal to ${n\mathbin{:}1\mathbin{:}j}$.
$[A,B]$ and $[A;B]$ are horizontal and vertical concatenation of matrices.
$0_{m\times n}$ and $1_{m\times n}$ are ${m\times n}$ matrices filled with $0$ and $1$.
$\0$ and $\1$ are 0 and 1 filled vectors of adequate size.
$\indicator{x}$ is the indicator function returning $1$ if $x$ is true and $0$ otherwise.
$\preceq$ is the component-wise less or equal for vectors and matrix inequality for matrices in ${\symmetric^{n}=\setbuild{X:\R^{n\times n}}{X'=X}}$.
${\defpos^{n}=\setbuild{X:\symmetric^{n}}{X\succ 0}}$, ${\semidefpos^{n}=\setbuild{X:\symmetric^{n}}{X\succeq 0}}$.
$I_{c}$ is the identity matrix of size $c$.
$A^{\dagger}$ is the generalized inverse ${{(A^{T}A)}^{-1}A^{T}}$.
$|A|$ is the determinant of $A$.
A vector $\vec{v}_{i}$, correspond to the $i$-th agent, and these vectors can be stacked in a vector $\vec{v}$.
$\random{a},\random{A},\randomvec{a}$ are random variables, scalar, matrix and vector respectively.
$\expectation{\random{x}}$ is the expected value of $\random{x}$, and $\probability{A \mid B}$ is the conditional probability of $A$ given condition $B$.
$\pdf{\cdot}$ is a probability density function. ${\diag(A_{1},\dots,A_{N})}$ corresponds to a block diagonal matrix.

\section{Problem Statement}\label{sec:PS}
In this section we present an algorithm to decompose a MPC problem based on the primal decomposition (\cite{BoydEtAl2015}), and a vulnerability that can be exploited by a malicious agent.
\subsection{Model Predictive Control}\label{ssec:MPC}
Let us consider a system composed of $M$ discrete-time linear time-invariant agents.
The dynamics of an agent $i$ is ruled by:
\begin{equation}
\begin{matrix}
  \label{eq:systems}
\vec{x}_{i}[k+1]&=&A_{i}\xik &+& B_{i}\uik\\
\vec{y}_{i}[k+1]&=&C_{i}\xik &&\\
\end{matrix}
\end{equation}
Each agent's input $\uik$ is constrained by
\begin{equation}
  \label{eq:u_local_constraints}
  \0\preceq\uik\preceq\vec{u}_{\max},
\end{equation}
where ${\umax:\R^{n_{u}\times 1}}$ and ${n_{u}=\card{\uik}}$.

All agents are coupled by a global input constraint
\begin{equation}
  \label{eq:global_constraint}
  \sum^{\nsubsystems}_{i=1}\Gamma_{i}\uik\preceq\umax,
\end{equation}
where \todo{$\Gamma_{i}:\semidefpos^{n_{u}}$}.

The overall system can be controlled by a \mpc{}, by predicting the dynamics of the system for a finite prediction horizon $N_{p}$ and computing the optimal input by solving the following optimization problem:
\begin{problem}{Global \mpc{} Problem.}\label{pb:GOP}
\begin{equation*}
\begin{matrix}
\underset{\useq}{\mathrm{minimize}}&\resizebox{0.35\textwidth}{!}{$\overbrace{\sum\limits^{M}_{i=1} \overbrace{\sum_{j=1}^{\predhorz} \norm{\mpcvec{v}[i][k+j][k]}^{2}_{Q_i}+\norm{\mpcvec{u}[i][k+j-1][k]}^{2}_{R_i}}^{\textstyle{} \obji}}^{\textstyle{} \globobj}$}\\
\mathrm{subject~ to}&~\eqref{eq:systems},\eqref{eq:u_local_constraints}\ \mathrm{and}\ \eqref{eq:global_constraint}
\left\}\small
\begin{aligned}
  &\forall i\in \{1:\nsubsystems\}\\
  &\forall j\in \{1:\predhorz\}
\end{aligned}\right.

\end{matrix}
% \label{eq:dmpcModLOP}
\end{equation*}
where ${Q_{i}:\semidefpos}$, ${R_{i}:\defpos}$, and $\vik$ is a control objective.
\todo{Reference tracking can be written as ${\vik=\wik-C_{i}\xik}$, where $\wik$ is a state reference.
Disturbance rejection can be written as ${\vik=C_{i}\xik}$.
}

$\optglobobj$ denotes the optimal value of the objective function of problem~\ref{pb:GOP}, and ${\optuseq}$ is the optimal control sequences for all agents.
The problem~\ref{pb:GOP} is solved at each time $k$, and the ${\optuikk}$ are applied in each respective $i$ subsystem, following a receding horizon strategy.
\end{problem}

Due to the monolithic aspect of problem~\ref{pb:GOP}, the prediction horizon and the number of agents, solving the problem can be computationally intensive. One can see that if it were not for the global constraints~\eqref{eq:global_constraint}, the problem would be easily decomposed in $M$ parts and solvable in parallel.

\subsection{Distributed Model Predictive Control}\label{ssec:dMPC}
&continue from here
The technique consists of decomposing the
\todo{\dmpc{},
}
\begin{subequations}
  \begin{equation}
    \left.
      \small
      \begin{aligned}
        J_{i}^{\star}(\thetaik)&=\underset{\uiseq}{\mathrm{minimize}} \obji\\
        \mathrm{s.t.} &\quad\eqref{eq:systems}\ \&\ \eqref{eq:upositive}\\
        &\quad\Gamma_{i}\uik\preceq\thetaik:\lambdaik\\
      \end{aligned}
    \right\}
    \small
    \begin{aligned}
      &\forall i\in \{1:\nsubsystems\}\\
      &\forall j\in \{1:\predhorz\}
    \end{aligned}
    \label{eq:DOP_local}
  \end{equation}
  \begin{equation}
    \small
    \begin{aligned}
      J^{\star}&=\underset{\thetaseq}{\mathrm{minimize}}\sum^{\nsubsystems}_{i=1} J^{\star}_i(\thetaik)\\
      \mathrm{s.t.} &\quad \sum_{i=1}^{\nsubsystems}\thetaik\preceq\umax
    \end{aligned}
    \label{eq:DOP_master}
  \end{equation}
\end{subequations}

@@

Where ${\set{H} = \setbuild{\vec{\theta}}{\sum_{i=1}^{M}\vec{\theta}_{i}\preceq \vec{u}_{\max}\ \&\&\ \vec{\theta}_{i}\succeq\0,i }}$.
\todo{The sum $\sum_{i=1}^{M}\vec{\theta}_{i}$ can also be represented by the matrix multiplication $I_{c}^{M}\vec{\theta}$.
}

@@

where ${I_{c}^{M}=\kron{\1_{M,1}}{I_{c}}}$, with ${c=N_{p}\pi_{\uik}}$.
Observe that ${I_{c}^{M} \pseudoinv{I_{c}^{M}}\vec{v}=\frac{\sum_{i=0}^{M}\vec{v}_{i}}{M}}$.

\no{\todo[delete probably]{
  But observe in~\eqref{eq:projectedSubgradient} that if ${\vec{\theta}_{i}\p=\0}$, ${(\vec{\lambda}_{i}^{\star} (\vec{\theta}_{i}\p)-I_{c}^{M} \pseudoinv{I_{c}^{M}}\vec{\lambda}^{\star} (\vec{\theta}\p))}$ needs to be negative so $\vec{\theta}_{i}\pplusone$ becomes negative, which is impossible \question{is it though?} since $\lambdaik\succeq\0$, from the KKT optimality conditions, seen in~\cite[Chapter 5]{BoydVandenberghe2004}.
}}

@@

\subsection{Local Problems --- Formal Analysis}\label{ssec:FA}
% \todo{As seen in \cite{NogueiraEtAl2021}},

@@

If we choose reference tracking as our, for instance, we have:
\begin{equation}
\small\begin{matrix*}[l]
 H_i&=&\mathcal{D}_i^T\bar{Q}_i\mathcal{D}_i+\bar{R}_i\\
\fik&=&\mathcal{D}_i^T\bar{Q}_i(\mathcal{M}_i\xik-\Wik)\\

@@

We stack in vectors $\Uik$ and $\Wik$
the input and setpoint predictions sequences ${\uiseq}$ and ${\wiseq}$.
${\mathcal{M}_{i}}$  and ${\mathcal{D}_{i}}$ are the prediction matrices of the \mpc{}.
$\bar{Q}_{i}$, $\bar{R}_{i}$, and $\Theta_{i}$ are block diagonal matrices built repeating $\predhorz$ times $Q_{i}$, $R_{i}$, and $\Gamma_i$ respectively.
\todo{
  Observe that $H_{i}:\defpos$, and that $\fik$ varies with $k$.
}

\todo{As seen in~\cite{BemporadEtAl2002} and in~\cite{BorrelliEtAl2017}, the solution for a constrained linear quadratic regulation problem yields in a \pwa{} function.
Similarly, we can get an explicit  solution for its dual variables $\vec{\lambda}_{i}$, which is also a \pwa{} function with respect to $\vec{\theta}_{i}$:}
\begin{equation}
  \begin{aligned}
    \label{eq:lambdafuntheta}
    \lambdai=
    \begin{cases}
      -P_{i}^{1}\vec{\theta}_{i}-\vec{s}_{i}^{1}[k]&\text{if}\ G_{1}[k]\thetai \preceq z_{1}[k] \\
      \qquad\quad \vdots&\qquad\quad \vdots\\
      -P_{i}^{N}\vec{\theta}_{i}-\vec{s}_{i}^{N}[k]&\text{if}\ G_{N}[k]\thetai \preceq z_{N}[k] \\
    \end{cases}
  \end{aligned}
\end{equation}
The halfspaces defined by the set of pairs $(G_{j}[k],z_{j}[k])$ represent a combination of active/inactive constraints of $\Theta_{i}\Uik\preceq\vec{\theta}_{i}$ for a given time $k$.

\begin{challenge}\label{ch:partition_unknown}
  It is important to note that since the actual state is unknown by the coordinator, it cannot anticipate the partition of the space for each agent.
\end{challenge}
\begin{challenge}\label{ch:zone_unknown}
 For the same reason, the values of $\vec{s}_{i}[k]$ are also unknown.
\end{challenge}

 As each constraint can be active or inactive, for a group of $n$ constraints, we can have at most $N=2^{n}$ different combinations of active/inactive sets and consequently $N$ regions (or zones).
 \begin{remark}\label{re:N_different_zones}
  Observe that depending on the constraints, some of these combinations can be redundant or even make the optimization infeasible.
 \end{remark}
 \begin{assumption}
   We assume that none of these cases in Remark~\ref{re:N_different_zones} can occur, and we have $N$ zones.
 \end{assumption}

@@

\subsection{Attack in DMPC scheme}\label{ssec:attack}

@@

\section{Simulation}
\begin{figure}[h]
  \centering
 \includegraphics[width=\columnwidth]{airtemp_roomI/__ErrorWX_command_normErrH.pdf}
  \caption{Air temperature in room I and the Frobenius norm of the decision variable $E_{I}[k]$ for different scenarios: nominal (N), selfish behavior without correction (S),
and selfish behavior with correction.}\label{fig:response3Scenarios}
\end{figure}
\begin{figure}[h]
  \centering
 \includegraphics[width=\columnwidth]{airtemp_roomI/control.pdf}
  \caption{Control applied in all rooms for the 3 different scenarios: nominal, selfish behavior without correction, and selfish behavior with correction (C).}\label{fig:response3Scenarios}
\end{figure}

% TODO(accacio): Verify
\begin{table}[t]
  \centering
  \caption{Comparison of costs $J_{i}$}\label{tab:costsGlobalLocal}
  \begin{tabular}[t]{cccc} \toprule
    Agent  & Nominal & Selfish & Selfish + Correction\\
    \midrule
    I & -11468 &  -9074 & -11468 \\
    II & -6536 &  -7151 & -6536 \\
    III & -3192 &  -4052 & -3192 \\
    IV & -2849 &  -3694 & -2849 \\
    Global & -24044 &  -23971 & -24044 \\

    \bottomrule
  \end{tabular}
\end{table}

@@

A conclusion section is not required. Although a conclusion may review
the main points of the paper, do not replicate the abstract as the
conclusion. A conclusion might elaborate on the importance of the work
or suggest applications and extensions.
